A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period

### 2021-01-12 analysis/paper/paper.Rmd
working on fixing typos

<!--burial study- network analysis-->
Burials analysis by archaeologists is an approach to understand past social structure through the study of the physical traces of mortuary practices. The material cultures and biological records of burial behaviors can represent the social ranking or identity of the deceased [@Binford1971; @Drennan2010; @Saxe1970]. Despite criticism that manipulation of burial rituals by the living can cause a disconnect between a person's status in life and their status represented by burial contexts [@Hodder1980; @Pearson1982], burials can still provide valuable information to infer past societies [@Chapman2003]. For example, mortuary practices, including burial forms, grave goods, and ritual behaviors, that structure the material configuration of burials can represent social relations between members of a community, allowing the inference of social structures [@Byrd1995; @Seikel2011]. Based on network science and graph theories, the recent development of social network analysis provides many new tools to visualize and analyze relationships for archaeological data [@Borgatti2009; @Brughmans2013]. Network science is the study of relational data where a phenomenon is conceptualized as a network through two steps, abstraction and representation, based on our specification in terms of network theories [@Brandes2013; @Collar2015]. That is, a network can be viewed as a patterned aggregation that includes individual elements (i.e. individual burials), pair-wise relationships (the dyads, for example burials with similar types and amounts of grave goods), and an overall structure showing global patterns represented in the data.

<!-- intro to networks -->
A social network is generally visualized as a graph consisting of a set of socially-relevant nodes/actors, connected by edges/ties representing one or more relations, such as friendship, collaborations, information flow, trade ties, or any other forms of connection of interest [@Marin2011; @Wasserman1994]. The ties can be classified into four major types, including similarities, social relations, interactions, and flows [@Borgatti2009]. In archaeology, actors can be people, groups, objects, places, or events, with ties built on similarity, proximity, or co-presence of material culture to create patterns reflecting influence, geographical distance, or affiliations in social groups [@Mills2017; @Brughmans2018; @Peeples2019]. For example, past trade can be conceptualized as a network of individual entities connected by shared similarity, the flow of goods, to represent their interactions [@Collar2015, pp. 4]. Similar concepts can be applied to a wide range of archaeological data with relational assumptions, such as burial contexts. Burial goods, especially high value goods, can reflect social practices in broader cultural contexts to represent personal wealth or social status from which we can infer social differentiation or complexity [@Gamble2002; @Janes2013]. Burials with the same prestige goods could indicate some underlying social relations where individuals share similar access to trade, exchange, and gifting networks according to their status [eg. @Coward2013, pp. 252]. This enables the exploration of the structure of the past social organization through the identification of the relationships among burials.

@@

<!-- archaeological specific case study -->
In this study, we explore the changes in social inequality in a pericolonial context at Kiwulan (Figure \@ref(fig:figure-KWL-map)), an Iron Age site in northeastern Taiwan, which covers the time before the European arrival, the presence of the Europeans, including the Spanish and the Dutch, in the 17th century, and finally the Chinese in the 19th century [@Chen2007]. We assume that the social changes at Kiwulan would be supported by chronological differences in the structure of burial networks. We examine burial networks before and after the foreign contacts, and test our hypothesis that increased social inequality can be observed in the network after the European arrival. We ask: (1) does the observed burial data after the start of European presence resemble the network of our model? (2) did European colonial activities in 17th century Taiwan result in an increased social inequality in an Indigenous society in ways that can be detected by burials networks? (3) what are the major variables affecting or forming the higher degree of popularity (e.g. a few nodes have more relationships with other nodes) in the European contact period that might hint at social heterogeneity during this period? By answering these questions, this study helps to expand the use of statistical modeling for burials in understanding the indirect effects of a colonial presence on Indigenous groups. This is important to understand the reactions of Kiwulan residents in the context of indirect impacts of European colonization.

@@

@@

Bayesian approach to Exponential Random Graph Models (ERGMs) is an effective way to examine relationships by incorporating prior information into networks to better understand dependencies of variables in complex networks and improve computational issues in ERGMs [@Caimo2017; @Lehmann2020]. Current network analysis used by archaeologist is mostly restricted to a single rational structure without consideration of interaction between network variables. Our use of Bayesian inference on ERGMs is the first application to archaeological data that can bring new insights to understand past social structures by characterizing network properties as a whole. ERGMs are an important family of statistical models for networks that allows direct modeling for the formation of edges, or ties, between nodes [@Robins2007]. The assumption is that possible ties in a network are random variables and dependent on actor variables or the presence or absence of other ties [@Robins2007]. In other words, networks in ERGMs are viewed as dependent variables, where network dependencies and the attributes of nodes/edges can influence the formation of a tie [@Snijders2011]. For example, nodes with similar attributes are more likely to form a relationship, such as friendship between people with the same hobby. Ties form a small structure in a network called a graph configuration, that describes the form of dependence, such as reciprocity (relationship between two actors), transitivity or clustering (relationship between two actors through a shared third actor), homophily (relationship between actors with a similar attribute), and popularity (actors have many relationships with others) [@Robins2007; @Snijders2006; @Morris2008]. Those configurations represent the structure or the property of a network and can be expressed by network statistics. By modeling those network statistics as direct functions of ties by specifying the forms of configurations, we can generate a distribution of random networks that represent our hypothesis-based model [@Morris2008; @Robins2007]. Such distribution consists of a large number of possible networks that enables statistical inference and comparison with an observed network [@Robins2007]. ERGMs helps us understand whether an observed network shows significantly more or less of a property of interest than the random networks generated from our model assumptions.

<!-- ERGM in archaeology, and limitations --> 
The use of ERGM in archaeology is first introduced by @Brughmans2014, who studied the Iron age settlement pattern based on 159 Iron Age sites in Southern Spain by modeling inter-settlement visibility networks and visual control. Although their models do not fit perfectly with the observed network of archaeological data, they proposed that ERGM is a promising method for exploration and hypothesis testing for social processes. Similarly, @Amati2019 reconstructed three different networks consisting of 15 sites (AD 100 to 400) in the Caribbean to explore their interaction mechanisms. These network models evaluated different hypotheses about the importance of three variables, proximity, inter-cultural items, and pottery types, on the presence of hub sites they observed. They found that the presence of hubs can be efficiently explained by multiple interdependent mechanisms instead of only one variable exclusively. However, those studies also point out some limitations of ERGM, such as sensitivity to missing data and less ability to handle uncertainty. In addition, it is difficult in ERGM to estimate model parameters and interpret the result due to intractable likelihood normalizing constants and model degeneracy [@Caimo2014; @Jin2013]. A normalizing constant is a function of the model parameter for making probability distributions integrate to one, which becomes harder to compute with larger set of networks [@Caimo2017bayesian]. This is also termed "doubly intractable" since both the likelihood normalizing constant and the marginal likelihood (the evidence of the posterior) are hard to derive [@Caimo2013; @Lyne2015]. Model degeneracy is another issue that probability models tend to overestimate a small number of extreme graphs by assigning too much weight, such as empty (all nodes unconnected) or complete graphs (all nodes connected) [@Schweinberger2011; @Caimo2014]. One solution to these limitations is available by implementing ERGMs in a Bayesian framework.

<!-- Bayesian inference for Exponential Random Graph Models --> 
Bayesian approaches to ERGMs are efficient computational tools for social network analysis because they incorporates prior information about the network configurations into the model and offers uncertainty quantification by evaluating the posterior distributions of the parameters associated with network configurations [@Caimo2015; @Nemmers2019]. In Bayesian analysis, posterior distribution is an updated probability distribution after combining the prior derived from previous data or assumptions and our data. The advantage that Bayesian modeling has over traditional ERGMs is the application of Markov chain Monte Carlo (MCMC) simulation using the approximate exchange algorithm [@Caimo2011]. MCMC avoids doubly-intractable computations by directly sampling from the not normalized part of the posterior, which alleviates the computational problems and gives a better convergence results. This enable us to deal with complicated dependence patterns with ease, providing better estimations for complex social network models with heterogeneous data [@Caimo2017; @Snijders2006]. By fitting an ERMG with an MCMC algorithm, a Bayesian approach generates posterior probabilities that incorporates our sample data and prior information through summary statistics from estimates of the ERGM parameters. Posterior probability estimates the effect of ERGM parameters by looking at the posterior mean and 95% credible intervals, which can replace p-values for assessing a null hypothesis [@Caimo2017bayesian]. The typical criteria for interpreting the posteriors is that an odds ratio greater than one means a positive effect of a parameter, while odds ratios less than one represents negative effect. In addition, Bayesian approaches are also useful to deal with missing data, which is often a problem leading to misinterpretation of networks, especially for archaeological studies. @Koskinen2010 shows that the effect of missing data can be reduced with Bayesian modeling that can predict, on the average, 80% of the ties when a third of data missing.

>>>>

<!--burial study- network analysis-->
Burial analysis by archaeologists is an approach to understand past social structures through the study of the physical traces of mortuary practices. The material cultures and biological records of burial behaviors can represent the social ranking or identities of the deceased [@Binford1971; @Drennan2010; @Saxe1970]. Despite criticism that manipulation of burial rituals by the living can cause a disconnect between a person's status in life and their status represented by burial contexts [@Hodder1980; @Pearson1982], burials can still provide valuable information to infer past societies [@Chapman2003]. For example, mortuary practices, including burial forms, grave goods, and ritual behaviors that structure the material configuration of burials, can represent social relations between members of a community, allowing the inference of social structures [@Byrd1995; @Seikel2011]. Based on network science and graph theory, the recent development of social network analysis provides many new tools to visualize and analyze relationships in archaeological data [@Borgatti2009; @Brughmans2013]. Network science is the study of relational data where a phenomenon is conceptualized as a network through two steps, abstraction and representation, based on our specification in terms of network theories [@Brandes2013; @Collar2015]. A network can be viewed as a patterned aggregation that includes individual elements (i.e. individual burials), pair-wise relationships (the dyads, for example burials with similar types and amounts of grave goods), and an overall structure showing global patterns represented in the data.

<!-- intro to networks -->
A social network is generally visualized as a graph consisting of a set of socially-relevant nodes/actors, connected by edges/ties representing one or more relations, such as friendship, collaborations, information flow, trade ties, or any other forms of connection of interest [@Marin2011; @Wasserman1994]. The ties can be classified into four major types, including similarities, social relations, interactions, and flows [@Borgatti2009]. In archaeology, actors can be people, groups, objects, places, or events, with ties built on similarity, proximity, or co-presence of material culture to create patterns reflecting influence, geographical distance, or affiliations in social groups [@Mills2017; @Brughmans2018; @Peeples2019]. For example, past trade can be conceptualized as a network of individual entities connected by shared similarities, the flow of goods, to represent their interactions [@Collar2015, pp. 4]. Similar concepts can be applied to a wide range of archaeological data with relational assumptions, such as burial contexts. Burial goods, especially high value goods, can reflect social practices in broader cultural contexts to represent personal wealth or social status from which we can infer social differentiation or complexity [@Gamble2002; @Janes2013]. Burials with the same prestige goods could indicate some underlying social relations where individuals share similar access to trade, exchange, and gifting networks according to their status [eg. @Coward2013, pp. 252]. This enables the exploration of the structure of the past social organization through the identification of the relationships among burials.

@@

<!-- archaeological specific case study -->
In this study, we explore the changes in social inequality in a pericolonial context at Kiwulan (Figure \@ref(fig:figure-KWL-map)), an Iron Age site in northeastern Taiwan, which covers the time before the European arrival, the presence of the Europeans, including the Spanish and the Dutch, in the 17th century, and finally the Chinese in the 19th century [@Chen2007]. We assume that social changes at Kiwulan would be supported by chronological differences in the structure of burial networks. We examine burial networks before and after the foreign contacts, and test our hypothesis that increased social inequality can be observed in the network after the European arrival. We ask: (1) did European colonial activities in 17th century Taiwan result in an increased social inequality in an Indigenous society in ways that can be detected by analysis of burial networks? (3) what are the major variables affecting or forming the higher degree of popularity (e.g. a few nodes have more relationships with other nodes) in the European contact period that might hint at social heterogeneity during this period? By answering these questions, this study helps to expand the use of statistical modeling for burials in understanding the indirect effects of a colonial presence on Indigenous groups. This is important to understand reactions to indirect impacts of European colonization.

@@

<!-- ERGM in archaeology, and limitations --> 
ERGMs first appeared in archaeology with @Brughmans2014, who studied  Iron age settlement patterns in Southern Spain by modeling inter-settlement visibility networks and visual control at 159 sites. Although their models do not fit perfectly with the observed network of archaeological data, they proposed that ERGM is a promising method for exploration and hypothesis testing for social processes. Similarly, @Amati2019 reconstructed three different networks consisting of 15 sites (AD 100 to 400) in the Caribbean to explore interaction mechanisms. These network models evaluated different hypotheses about the importance of three variables, proximity, inter-cultural items, and pottery types, on the presence of hub sites they observed. They found that the presence of hubs can be efficiently explained by multiple interdependent mechanisms instead of only one variable exclusively. However, those studies also point out some limitations of ERGM, such as sensitivity to missing data and less ability to handle uncertainty. In addition, it is difficult in ERGM to estimate model parameters and interpret the result due to intractable likelihood normalizing constants and model degeneracy [@Caimo2014; @Jin2013]. A normalizing constant is a function of the model parameter for making probability distributions integrate to one, which becomes harder to compute with larger set of networks [@Caimo2017bayesian]. This is also termed "doubly intractable" since both the likelihood normalizing constant and the marginal likelihood (the evidence of the posterior) are hard to derive [@Caimo2013; @Lyne2015]. Model degeneracy is another issue that probability models tend to overestimate a small number of extreme graphs by assigning too much weight, such as empty (all nodes unconnected) or complete graphs (all nodes connected) [@Schweinberger2011; @Caimo2014]. One solution to these limitations is available by implementing ERGMs in a Bayesian framework.

@@

Bayesian approaches to Exponential Random Graph Models (ERGMs) are effective ways to examine relationships by incorporating prior information into networks to better understand dependencies of variables in complex networks and improve computational issues in ERGMs [@Caimo2017; @Lehmann2020]. Current approaches to network analysis used by archaeologists are mostly restricted to a single rational structure without consideration of interaction between network variables. Our use of Bayesian inference on ERGMs is the first application to archaeological data that can bring new insights to understand past social structures by characterizing network properties as a whole. ERGMs are an important family of statistical models for networks that allows direct modeling for the formation of edges, or ties, between nodes [@Robins2007]. The assumption is that possible ties in a network are random variables and dependent on actor variables or the presence or absence of other ties [@Robins2007]. In other words, networks in ERGMs are viewed as dependent variables, where network dependencies and the attributes of nodes/edges can influence the formation of a tie [@Snijders2011]. For example, nodes with similar attributes are more likely to form a relationship, such as friendship between people with the same hobby. Ties form a small structure in a network called a graph configuration, that describes the form of dependence, such as reciprocity (relationship between two actors), transitivity or clustering (relationship between two actors through a shared third actor), homophily (relationship between actors with a similar attribute), and popularity (actors have many relationships with others) [@Robins2007; @Snijders2006; @Morris2008]. Those configurations represent the structure or the property of a network and can be expressed by network statistics. By modeling those network statistics as direct functions of ties by specifying the forms of configurations, we can generate a distribution of random networks that represent our hypothesis-based model [@Morris2008; @Robins2007]. Such a distribution consists of a large number of possible networks that enables statistical inference and comparison with an observed network [@Robins2007]. ERGMs helps us understand whether an observed network shows significantly more or less of a property of interest than the random networks generated from our model assumptions.

<!-- Bayesian inference for Exponential Random Graph Models --> 
Bayesian approaches to ERGMs are efficient computational tools for social network analysis because they incorporate prior information about the network configurations into the model and offer uncertainty quantification by evaluating the posterior distributions of the parameters associated with network configurations [@Caimo2015; @Nemmers2019]. In Bayesian analysis, posterior distribution is an updated probability distribution after combining the prior derived from previous data or assumptions and our data. The advantage that Bayesian modeling has over traditional ERGMs is the application of Markov chain Monte Carlo (MCMC) simulation using the approximate exchange algorithm [@Caimo2011]. MCMC avoids doubly-intractable computations by directly sampling from the not normalized part of the posterior, which alleviates the computational problems and gives a better convergence results. This enable us to deal with complicated dependence patterns with ease, providing better estimations for complex social network models with heterogeneous data [@Caimo2017; @Snijders2006]. By fitting an ERMG with an MCMC algorithm, a Bayesian approach generates posterior probabilities that incorporates our sample data and prior information through summary statistics from estimates of the ERGM parameters. Posterior probability estimates the effect of ERGM parameters by looking at the posterior mean and 95% credible intervals, which can replace p-values for assessing a null hypothesis [@Caimo2017bayesian]. The typical criteria for interpreting the posteriors is that an odds ratio greater than one means a positive effect of a parameter, while odds ratios less than one represents negative effect. In addition, Bayesian approaches are also useful to deal with missing data, which is often a problem leading to misinterpretation of networks, especially for archaeological studies. @Koskinen2010 shows that the effect of missing data can be reduced with Bayesian modeling that can predict, on the average, 80% of the ties when a third of data missing.
